---
title: "T3 code"
output: html_document
---

# T3: Trip Distribution

**To run this code on a different computer, command/control + f `file_path` and change all the file paths**

# Load libraries

```{r, message = FALSE, warning = FALSE}
# Installing omxr
# install.packages("devtools")
# devtools::install_github("gregmacfarlane/omxr")

# Installing chorddiag
# devtools::install_github("mattflor/chorddiag")

library(tidyverse)
library(sf)
library(DBI)
library(here)
library(leaflet)
library(maptiles)
library(tidyterra)
library(knitr)
library(tigris)
library(omxr)
library(scales)
library(cowplot)
library(RColorBrewer)
library(chorddiag)
library(units)

sf::sf_use_s2(FALSE)
```

# Load trip matrices

### Base model

```{r, message = FALSE, warning = FALSE}
file_path <- "/Users/paigelee/Documents/SES 5394/T3/pa_trips_all_base_2.omx"

base_pa <- read_all_omx(file_path, c("hb", "nhb")) |>
  mutate(total_hh_trips = hb + nhb)

head(base_pa) |>
  kable(digits = 2)
```

### Alternative model

```{r, message = FALSE, warning = FALSE}
file_path <- "/Users/paigelee/Documents/SES 5394/T3/pa_trips_all_alt_2.omx"

alt_pa <- read_all_omx(file_path, c("hb", "nhb")) |>
  mutate(total_hh_trips = hb + nhb)

head(alt_pa) |>
  kable(digits = 2)
```

# Visualization

### Map of the TAZ IDs in our model

```{r, message = FALSE, warning = FALSE}
file_path = "/Users/paigelee/Documents/SES 5394/T3/CTPS_TDM23_TAZ_2017g_v202303.shp"

TAZs <- st_read(file_path, quiet = TRUE) |>
  st_transform("WGS84")

leaflet(TAZs) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addPolygons(weight = 2,
              color = "orange",
              fillColor = "orange",
              fillOpacity = 0.1,
              highlightOptions = highlightOptions(weight = 3,
                                                  fillOpacity = 0.5),
              label = TAZs$taz_id)
```

### Removing misnumbered TAZ IDs in RI and NH

```{r, message = FALSE, warning = FALSE}
TAZs <- TAZs |>
  filter(state == "MA",
         mpo != "BRPC",
         mpo != "FRCOG")

base_pa <- base_pa |>
  filter(origin %in% TAZs$taz_id,
         destination %in% TAZs$taz_id)

alt_pa <- alt_pa |>
  filter(origin %in% TAZs$taz_id,
         destination %in% TAZs$taz_id)

leaflet(TAZs) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addPolygons(weight = 2,
              color = "orange",
              fillColor = "orange",
              fillOpacity = 0.1,
              highlightOptions = highlightOptions(weight = 3,
                                                  fillOpacity = 0.5),
              label = TAZs$taz_id)
```

# Mapping intrazonal trips

Intrazonal trips begin and end in the same zone

```{r, message = FALSE, warning = FALSE}
base_productions <- base_pa |>
  group_by(origin) |>
  summarize(productions = sum(total_hh_trips)) |>
  rename(taz_id = origin)

base_attractions <- base_pa |>
  group_by(destination) |>
  summarize(attractions = sum(total_hh_trips)) |>
  rename(taz_id = destination)

base_intrazonal <- base_pa |>
  filter(origin == destination) |>
  select(origin, total_hh_trips) |>
  rename(taz_id = origin,
         intrazonal = total_hh_trips) |>
  full_join(base_productions) |>
  full_join(base_attractions) |>
  filter(productions + attractions > 0) |>
  mutate(pct_intra_attr = intrazonal / attractions,
         pct_intra_prod = intrazonal / productions) 

base_intrazonal <- TAZs |>
  inner_join(base_intrazonal)

base_map <- get_tiles(base_intrazonal,
                      provider = "CartoDB.Positron",
                      zoom = 8,
                      crop = TRUE)
```

### Map of the number of intrazonal trips for the base model

```{r, message = FALSE, warning = FALSE}
ggplot(base_intrazonal) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(color = NA,
          alpha = 0.7,
          aes(fill = intrazonal)) +
  scale_fill_viridis_c(transform = "log",
                       breaks = breaks <- 10^seq(-3, 4, by=1),
                       labels = c(formatC(breaks[1:3], 
                                        format = "f",
                                        digits = 3),
                                  formatC(breaks[4:8],
                                          format = "d",
                                        big.mark = " ")),
                       name = "Number of\nintrazonal trips") +
  theme_void()
```

### Map of the percentage of all productions that are for intrazonal trips for the base model

```{r, message = FALSE, warning = FALSE}
ggplot(base_intrazonal) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(color = NA,
          alpha = 0.7,
          aes(fill = pct_intra_prod)) +
  scale_fill_viridis_c(name = "Intrazonal trips\n(% of productions)",
                       breaks = breaks <- seq(0, 0.7, by=0.1),
                       labels = paste0(round(breaks*100),"%")) +
  theme_void()
```

### Map of the percentage of all attractions that are for intrazonal trips for the base model

```{r, message = FALSE, warning = FALSE}
ggplot(base_intrazonal) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(color = NA,
          alpha = 0.7,
          aes(fill = pct_intra_attr)) +
  scale_fill_viridis_c(name = "Intrazonal trips\n(% of attractions)",
                       breaks = breaks <- seq(0, 0.9, by=0.1),
                       labels = paste0(round(breaks*100),"%")) +
  theme_void()
```

# Desire lines

A desire line is a straight line from the trip's origin (production) to its destination (attraction). We use the zone centroids as endpoints of the desire lines.

### Plotting zone centroids

```{r, message = FALSE, warning = FALSE}
TAZ_pts <- st_centroid(TAZs)

leaflet(TAZ_pts) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addCircleMarkers(weight = 2,
             color = NA,
             fillColor = "orange",
             radius = 3,
             fillOpacity = 0.6,
             label = TAZ_pts$taz_id)
```

### Function to create a set of desire lines 

```{r, message = FALSE, warning = FALSE}
make_desire_lines <- function(od_df,
                              points,
                              origin_column,
                              destination_column,
                              trips_column,
                              taz_id_column) {
  
  od_df <- od_df |>
    ungroup() |>
    rename(origin = all_of(origin_column),
           destination = all_of(destination_column),
           trips = all_of(trips_column)) |>
    select(origin, destination, trips) |>
    filter(origin != destination)
  
  points <- points |>
    rename(taz_id = all_of(taz_id_column))
  
  desire_line_data <- od_df |>
    mutate(id = seq(1, nrow(od_df))) |>
    pivot_longer(cols = c(origin, destination),
                 names_to = "o_d",
                 values_to = "taz_id") 

  desire_lines <- points |>
    right_join(desire_line_data) |>
    group_by(id) |>
    summarize(trips = mean(trips),
              origin_taz = first(taz_id),
              destin_taz = last(taz_id)) |>
    ungroup() |>
    select(origin_taz, destin_taz, trips) |>
    st_cast("LINESTRING")
  
  desire_lines

}
```

### Defining a smaller focus area 

We want to only generate a set of desire lines for a smaller number of zones in the vicinity of where we made changes. We want to identify all TAZ centroids within 1 mile of the TAZ where we made changes for our alternative scenario 

```{r, message = FALSE, warning = FALSE}
# Manually defined list of TAZ centroids where we made changes for our alternative model
taz_list <- c(643, 642, 644, 691, 695, 691, 238)

focus_area <- TAZ_pts[TAZ_pts$taz_id %in% taz_list,] |>
  st_transform(2249) |>
  st_buffer(dist = 5280)

focus_area_pts <- TAZ_pts |>
  st_transform(2249) |>
  st_filter(focus_area) |>
  st_transform("WGS84")

leaflet(focus_area_pts) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addCircleMarkers(weight = 2,
             color = NA,
             fillColor = "orange",
             radius = 5,
             fillOpacity = 0.6,
             label = focus_area_pts$taz_id)
```

### Generating desire lines for our smaller focus area

```{r, message = FALSE, warning = FALSE}
focus_area_base_pa <- base_pa |>
  filter(origin %in% focus_area_pts$taz_id,
         destination %in% focus_area_pts$taz_id)

desire_lines_base_focus <- make_desire_lines(od_df = focus_area_base_pa,
                                             points = focus_area_pts,
                                             origin_column = "origin",
                                             destination_column = "destination",
                                             trips_column = "total_hh_trips",
                                             taz_id_column = "taz_id")

base_map <- get_tiles(desire_lines_base_focus,
                      provider = "CartoDB.Positron",
                      zoom = 14,
                      crop = TRUE)

ggplot(desire_lines_base_focus) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(aes(linewidth = trips),
          alpha = 0.1,
          color = "coral") +
  theme_void()
```

### Aggregating by towns

```{r, message = FALSE, warning = FALSE}
boston_town_pts <- TAZ_pts |>
  filter(mpo == "BRMPO") |>
  group_by(town) |>
  summarise(n_zones = n()) |>
  st_centroid()

TAZ_town_codes <- st_drop_geometry(TAZ_pts) |>
  filter(mpo == "BRMPO") |>
  select(taz_id, town) 

base_pa_towns <- base_pa |>
  rename(taz_id = origin) |>
  left_join(TAZ_town_codes) |>
  rename(origin_town = town,
         origin = taz_id) |>
  rename(taz_id = destination) |>
  left_join(TAZ_town_codes) |>
  rename(destin_town = town,
         destination = taz_id) |>
  group_by(origin_town, destin_town) |>
  summarize(trips = sum(total_hh_trips)) |>
  filter(!is.na(origin_town),
         !is.na(destin_town))
```

### Generating desire lines among towns

```{r, message = FALSE, warning = FALSE}
desire_lines_base_towns <- make_desire_lines(od_df = base_pa_towns,
                                             points = boston_town_pts,
                                             origin_column = "origin_town",
                                             destination_column = "destin_town",
                                             trips_column = "trips",
                                             taz_id_column = "town")

base_map <- get_tiles(desire_lines_base_towns,
                      provider = "CartoDB.Positron",
                      zoom = 10,
                      crop = TRUE)

ggplot(desire_lines_base_towns) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(aes(linewidth = trips),
          alpha = 0.01,
          color = "coral") +
  theme_void()
```

### Limiting the map of desire lines among towns to a smaller number of towns

```{r, message = FALSE, warning = FALSE}
fewer_towns <- c("CAMBRIDGE",
                            "BOSTON",
                            "SOMERVILLE",
                            "MEDFORD",
                            "EVERETT",
                            "MALDEN")

base_fewer_pa_towns <- base_pa_towns |>
  filter(origin_town %in% fewer_towns,
         destin_town %in% fewer_towns)

desire_fewer_base_towns <- make_desire_lines(od_df = base_fewer_pa_towns,
                                             points = boston_town_pts,
                                             origin_column = "origin_town",
                                             destination_column = "destin_town",
                                             trips_column = "trips",
                                             taz_id_column = "town")
```

```{r, message = FALSE, warning = FALSE}
base_map <- get_tiles(desire_fewer_base_towns,
                      provider = "CartoDB.Positron",
                      zoom = 13,
                      crop = TRUE)

ggplot(desire_fewer_base_towns) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(aes(linewidth = trips),
          alpha = 0.5,
          color = "coral") +
  theme_void()
```

### Aggregating by MPO areas (larger than towns)

```{r, message = FALSE, warning = FALSE}
mpo_pts <- TAZ_pts |>
  group_by(mpo) |>
  summarise(n_zones = n()) |>
  st_centroid()

TAZ_mpo_codes <- st_drop_geometry(TAZ_pts) |>
  select(taz_id, mpo) 

base_pa_mpos <- base_pa |>
  rename(taz_id = origin) |>
  left_join(TAZ_mpo_codes) |>
  rename(origin_mpo = mpo,
         origin = taz_id) |>
  rename(taz_id = destination) |>
  left_join(TAZ_mpo_codes) |>
  rename(destin_mpo = mpo,
         destination = taz_id) |>
  group_by(origin_mpo, destin_mpo) |>
  summarize(trips = sum(total_hh_trips)) |>
  filter(!is.na(origin_mpo),
         !is.na(destin_mpo))
```

### Generating desire lines among MPOs

```{r, message = FALSE, warning = FALSE}
desire_base_mpos <- make_desire_lines(od_df = base_pa_mpos,
                                      points = mpo_pts,
                                      origin_column = "origin_mpo",
                                      destination_column = "destin_mpo",
                                      trips_column = "trips",
                                      taz_id_column = "mpo")

base_map <- get_tiles(desire_base_mpos,
                      provider = "CartoDB.Positron",
                      zoom = 8,
                      crop = TRUE)

ggplot(desire_base_mpos) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(aes(linewidth = trips),
          alpha = 0.5,
          color = "coral") +
  theme_void()
```

# Chord diagrams

We can use chord diagrams to visualize the origin-destination (production-attraction) flows without a map

### Chord diagram among the smaller selected number of towns

```{r, message = FALSE, warning = FALSE}
mat <- matrix(base_fewer_pa_towns$trips, 
              nrow = 6, 
              ncol = 6, 
              byrow = TRUE) 

chorddiag(mat, groupNames = base_fewer_pa_towns$destin_town[1:6])
```

# Comparing the base vs. alternative scenarios

### Comparing the numbers of trips

Setting `tolerance <- 10` means that we'll recognize a change as having at least a change of 10 trips between a pair of zones between the base vs. alternative models

```{r, message = FALSE, warning = FALSE}
tolerance <- 10

base_pa <- base_pa |>
  mutate(total_hh_trips_base = hb + nhb) |>
  select(origin, destination, total_hh_trips_base)

alt_pa <- alt_pa |>
  mutate(total_hh_trips_alt = hb + nhb) |>
  select(origin, destination, total_hh_trips_alt)

comparison <- inner_join(base_pa, alt_pa) |>
  mutate(difference = total_hh_trips_alt - total_hh_trips_base) |>
  filter(abs(difference) > tolerance)

head(comparison) |>
  kable(digits = 1)
```

### Mapping the locations impacted by the change

```{r, message = FALSE, warning = FALSE}
changed_TAZ_list <- unique(c(comparison$origin, comparison$destination))

lost_p <- comparison |>
  filter(difference < 0) |>
  group_by(origin) |>
  summarise(lost_p = -1 * sum(difference)) |>
  rename(taz_id = origin)

lost_a <- comparison |>
  filter(difference < 0) |>
  group_by(destination) |>
  summarise(lost_a = -1 * sum(difference)) |>
  rename(taz_id = destination)

gain_p <- comparison |>
  filter(difference > 0) |>
  group_by(origin) |>
  summarise(gain_p = sum(difference)) |>
  rename(taz_id = origin)

gain_a <- comparison |>
  filter(difference > 0) |>
  group_by(destination) |>
  summarise(gain_a = sum(difference)) |>
  rename(taz_id = destination)

comparison_summary <- full_join(gain_a,
                                gain_p) |>
  full_join(lost_a) |>
  full_join(lost_p) |>
  replace_na(list(lost_a = 0,
                  lost_p = 0, 
                  gain_a = 0,
                  gain_p = 0))

changed_TAZs <- TAZs |>
  right_join(comparison_summary)

change_labels <- paste0("Gained ",
                        round(changed_TAZs$gain_a),
                        " trips to here<br/>",
                        "Lost ",
                        round(changed_TAZs$lost_a),
                        " trips to here<br/>",
                        "Gained ",
                        round(changed_TAZs$gain_p),
                        " trips from here<br/>",
                        "Lost ",
                        round(changed_TAZs$lost_p),
                        " trips from here<br/>") |>
  lapply(htmltools::HTML)

leaflet(changed_TAZs) |>
  addProviderTiles(provider = "CartoDB.Positron") |>
  addPolygons(weight = 2,
              color = "orange",
              fillColor = "orange",
              fillOpacity = 0.1,
              highlightOptions = highlightOptions(weight = 3,
                                                  fillOpacity = 0.5),
              label = change_labels)
```

### Separating intrazonal vs. interzonal trips

```{r, message = FALSE, warning = FALSE}
intrazonal_trips <- comparison |>
  filter(origin == destination) |>
  rename(taz_id = origin) |>
  select(-destination)

interzonal_trips <- comparison |>
  filter(origin != destination) 
```

### Distribution of changes in intrazonal trips

```{r, message = FALSE, warning = FALSE}
ggplot(intrazonal_trips) +
  geom_histogram(aes(x = difference),
                 binwidth = 10,
                 fill = "orange",
                 alpha = 0.5,
                 color = "gray") +
  scale_y_continuous(breaks = seq(0, 12, by=1),
                     name = "Number of zones (among zones with\na differnce of at least +/- 10 trips)") +
  scale_x_continuous(breaks = seq(0, 600, by = 50),
                     name = "Difference in number of intrazonal trips") +
  theme_minimal()
```

```{r, message = FALSE, warning = FALSE}
intrazonal_locs <- changed_TAZs |>
  right_join(intrazonal_trips)

base_map <- get_tiles(intrazonal_locs,
                      provider = "CartoDB.Positron",
                      zoom = 13,
                      crop = TRUE)

# Fix transformation issue by filtering for non-zero positive values
ggplot(intrazonal_locs |> filter(difference > 10)) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(aes(fill = difference),
          color = "gray",
          alpha = 0.7) +
  scale_fill_viridis_c(
    trans = "log10",
    breaks = c(10 * 2^seq(0, 5)),
    direction = -1,
    name = "Increase in\nintrazonal\ntrips"
  ) +
  theme_void()
```

### Generating desire lines to visualize changes in demand for intrazonal travel

```{r, message = FALSE, warning = FALSE}
changed_TAZ_points <- changed_TAZs |>
  st_centroid() |>
  select(taz_id)

desire_line_data <- interzonal_trips |>
  mutate(id = seq(1, nrow(interzonal_trips))) |>
  pivot_longer(cols = c(origin, destination),
               names_to = "o_d",
               values_to = "taz_id") 

desire_line_change <- changed_TAZ_points |>
  right_join(desire_line_data) |>
  group_by(id) |>
  summarize(difference = mean(difference),
            origin_taz = first(taz_id),
            destin_taz = last(taz_id)) |>
  ungroup() |>
  select(difference, origin_taz, destin_taz) |>
  st_cast("LINESTRING")

base_map <- get_tiles(desire_line_change,
                      provider = "CartoDB.DarkMatter",
                      zoom = 12,
                      crop = TRUE)

ggplot(desire_line_change) + 
  geom_spatraster_rgb(data = base_map) +
  geom_sf(aes(color = difference)) +
  scale_color_gradient2(low = muted("red"), 
                        mid = "white",
                        high = muted("blue"),
                        midpoint = 0) +
  theme_void()
```

### Improving desire lines to visualize changes in demand for intrazonal travel

We separated gains (blue) and losses (red), used line thickness to indicate the magnitude of change, and increased the transparency of the lines

```{r, message = FALSE, warning = FALSE}
desire_line_gain <- desire_line_change |>
  filter(difference > 0)

desire_line_loss <- desire_line_change |>
  filter(difference < 0) |>
  mutate(difference = -1 * difference)

base_map <- get_tiles(desire_line_change,
                      provider = "CartoDB.Positron",
                      zoom = 12,
                      crop = TRUE)

ggplot() + 
  geom_spatraster_rgb(data = base_map) +
  geom_sf(data = desire_line_loss,
          alpha = 0.2,
          aes(linewidth = difference,
              color = "Reduced demand")) +
  geom_sf(data = desire_line_gain,
          alpha = 0.2,
          aes(linewidth = difference,
              color = "Increased demand")) +
  scale_linewidth(name = "Magnitude of difference\n(number of trips)") +
  scale_color_manual(name = "Direction of difference",
                     values = c(muted("blue"), muted("red"))) +
  guides(color = guide_legend(override.aes = list(linewidth = 2,
                                                  alpha = 0.5))) +
  theme_void()
```

### Plotting gains and losses on separate maps

```{r, message = FALSE, warning = FALSE}
gain_map <- ggplot() + 
  geom_spatraster_rgb(data = base_map) +
  geom_sf(data = desire_line_gain,
          alpha = 0.15,
          color = "orange",
          aes(linewidth = difference)) +
  scale_linewidth(name = "Magnitude of difference",
                  limits = c(0,500),
                  breaks = breaks <- seq(100, 500, by = 100),
                  labels = paste0(breaks, " trips")) +
  theme_void() 

loss_map <- ggplot() + 
  geom_spatraster_rgb(data = base_map) +
  geom_sf(data = desire_line_loss,
          alpha = 0.15,
          color = "orange",
          aes(linewidth = difference)) +
  scale_linewidth(name = "Magnitude of difference",
                  limits = c(0,500),
                  breaks = breaks <- seq(100, 500, by = 100),
                  labels = paste0(breaks, " trips")) +
  theme_void() 

legend <- get_legend(loss_map)

plot_grid(gain_map + theme(legend.position = "none"), 
          loss_map + theme(legend.position = "none"), 
          legend,
          nrow = 1,
          labels = c("Trip increases",
                     "Trip decreases",
                     ""),
          label_size = 10,
          label_y = 0.8,
          label_x = -0.12)
```

### Identifying the locations with the largest changes

```{r, message = FALSE, warning = FALSE}
most_gained_a <- comparison_summary$taz_id[
  comparison_summary$gain_a == max(comparison_summary$gain_a)]

most_gained_p <- comparison_summary$taz_id[
  comparison_summary$gain_p == max(comparison_summary$gain_p)]

most_lost_a <- comparison_summary$taz_id[
  comparison_summary$lost_a == max(comparison_summary$lost_a)]

most_lost_p <- comparison_summary$taz_id[
  comparison_summary$lost_p == max(comparison_summary$lost_p)]

tibble(Change = c("Greatest production gain",
                  "Greatest attraction gain",
                  "Greatest production loss",
                  "Greatest attraction loss"),
       TAZ = c(most_gained_p,
               most_gained_a,
               most_lost_p,
               most_lost_a),
       Magnitude = c(max(comparison_summary$gain_p),
                     max(comparison_summary$gain_a),
                     max(comparison_summary$lost_p),
                     max(comparison_summary$lost_a))) |>
  kable(digits = 1)
```

### Further examining the zones with the largest changes

TAZ IDs with the largest changes: 691 and 643

```{r, message = FALSE, warning = FALSE}
# Further examining TAZ ID 691

compare_taz_691 <- comparison |>
  filter(origin == 691 | destination == 691) |>
  mutate(taz_id = ifelse(origin == 691, destination, origin)) |>
  group_by(taz_id) |>
  summarise(difference = sum(difference)) 

taz_691_zones <- changed_TAZs |>
  right_join(compare_taz_691) |>
  select(difference)

base_map <- get_tiles(taz_691_zones,
                      provider = "CartoDB.Positron",
                      zoom = 12,
                      crop = TRUE)

ggplot(taz_691_zones) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(color = "gray",
          aes(fill = difference)) +
  scale_fill_gradient2(low = muted("red"),
                       mid = "white",
                       high = muted("blue"),
                       midpoint = 0) +
  theme_void()
```

```{r, message = FALSE, warning = FALSE}
# Further examining TAZ ID 643

compare_taz_643 <- comparison |>
  filter(origin == 643 | destination == 643) |>
  mutate(taz_id = ifelse(origin == 643, destination, origin)) |>
  group_by(taz_id) |>
  summarise(difference = sum(difference)) 

taz_643_zones <- changed_TAZs |>
  right_join(compare_taz_643) |>
  select(difference)

base_map <- get_tiles(taz_643_zones,
                      provider = "CartoDB.Positron",
                      zoom = 12,
                      crop = TRUE)

ggplot(taz_643_zones) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(color = "gray",
          aes(fill = difference)) +
  scale_fill_gradient2(low = muted("red"),
                       mid = "white",
                       high = muted("blue"),
                       midpoint = 0) +
  theme_void()
```



