---
title: "T5 code"
output: html_document
---

# Load libraries

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(sf)
library(here)
library(maptiles)
library(tidyterra)
library(knitr)
library(tigris)
```

# Load the highway network

```{r, message = FALSE, warning = FALSE}
file_path = "/Users/paigelee/Documents/SES 5394/T5/hwy.geojson"

all_roads <- st_read(file_path, quiet = TRUE)
```

```{r, message = FALSE, warning = FALSE}
ggplot(all_roads) +
  geom_sf() +
  theme_void()
```

# Load link volumes

Link volumes from the trip assignment step are in the following four .csv files:

* flows_am.csv (AM peak period)
* flows_pm.csv (PM peak period)
* flows_md.csv (mid-day off-peak period)
* flows_nt.csv (nighttime off-peak period)

```{r, message = FALSE, warning = FALSE}
# AM peak period

# Base
file_path = "/Users/paigelee/Documents/SES 5394/T5/flows_am_base.csv"
am_hwy_flows_base <- read_csv(file_path, show_col_types = FALSE) |>
  rename(ID = ID1,
         base_flow = Tot_Flow_PCE) |>
  select(ID, base_flow)

# Alt
file_path = "/Users/paigelee/Documents/SES 5394/T5/flows_am_alt.csv"

am_hwy_flows_alt <- read_csv(file_path, show_col_types = FALSE) |>
  rename(ID = ID1,
         alt_flow = Tot_Flow_PCE) |>
  select(ID, alt_flow)
```

```{r, message = FALSE, warning = FALSE}
# PM peak period

# Base
file_path = "/Users/paigelee/Documents/SES 5394/T5/flows_pm_base.csv"
pm_hwy_flows_base <- read_csv(file_path, show_col_types = FALSE) |>
  rename(ID = ID1,
         base_flow = Tot_Flow_PCE) |>
  select(ID, base_flow)

# Alt
file_path = "/Users/paigelee/Documents/SES 5394/T5/flows_pm_alt.csv"

pm_hwy_flows_alt <- read_csv(file_path, show_col_types = FALSE) |>
  rename(ID = ID1,
         alt_flow = Tot_Flow_PCE) |>
  select(ID, alt_flow)
```

```{r, message = FALSE, warning = FALSE}
# Mid-day off-peak period

# Base
file_path = "/Users/paigelee/Documents/SES 5394/T5/flows_md_base.csv"
md_hwy_flows_base <- read_csv(file_path, show_col_types = FALSE) |>
  rename(ID = ID1,
         base_flow = Tot_Flow_PCE) |>
  select(ID, base_flow)

# Alt
file_path = "/Users/paigelee/Documents/SES 5394/T5/flows_md_alt.csv"

md_hwy_flows_alt <- read_csv(file_path, show_col_types = FALSE) |>
  rename(ID = ID1,
         alt_flow = Tot_Flow_PCE) |>
  select(ID, alt_flow)
```

```{r, message = FALSE, warning = FALSE}
# Nighttime off-peak period

# Base
file_path = "/Users/paigelee/Documents/SES 5394/T5/flows_nt_base.csv"
nt_hwy_flows_base <- read_csv(file_path, show_col_types = FALSE) |>
  rename(ID = ID1,
         base_flow = Tot_Flow_PCE) |>
  select(ID, base_flow)

# Alt
file_path = "/Users/paigelee/Documents/SES 5394/T5/flows_nt_alt.csv"

nt_hwy_flows_alt <- read_csv(file_path, show_col_types = FALSE) |>
  rename(ID = ID1,
         alt_flow = Tot_Flow_PCE) |>
  select(ID, alt_flow)
```

# Join base and alt link volumes for each time of day

```{r, message = FALSE, warning = FALSE}
# AM peak period

link_vols_compare_am <- inner_join(all_roads, am_hwy_flows_base) |>
  left_join(am_hwy_flows_alt) |>
  replace_na(list(alt_flow = 0)) |>
  mutate(flow_diff = alt_flow - base_flow) 
```

```{r, message = FALSE, warning = FALSE}
# PM peak period

link_vols_compare_pm <- inner_join(all_roads, pm_hwy_flows_base) |>
  left_join(pm_hwy_flows_alt) |>
  replace_na(list(alt_flow = 0)) |>
  mutate(flow_diff = alt_flow - base_flow) 
```

```{r, message = FALSE, warning = FALSE}
# Mid-day off-peak period

link_vols_compare_md <- inner_join(all_roads, md_hwy_flows_base) |>
  left_join(md_hwy_flows_alt) |>
  replace_na(list(alt_flow = 0)) |>
  mutate(flow_diff = alt_flow - base_flow) 
```

```{r, message = FALSE, warning = FALSE}
# Nighttime off-peak period

link_vols_compare_nt <- inner_join(all_roads, nt_hwy_flows_base) |>
  left_join(nt_hwy_flows_alt) |>
  replace_na(list(alt_flow = 0)) |>
  mutate(flow_diff = alt_flow - base_flow) 
```

# Filtering to only include the area near where we made changes

Professor Voulgaris filtered to only include roads within 5 kilometers of a bridge she removed and provided the coordinates of that bridge

In our case, we want to filter the data to only include roads within 5 kilometers of Harvard and within 15 kilometers of MIT.

```{r, message = FALSE, warning = FALSE}
# Create buffers around Harvard and MIT
harvard_coords <- tibble(lat = 42.373611, lon = -71.109733) |>  # Harvard 
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>  # WGS84
  st_transform(26986) |>  # Massachusetts state plane (meters)
  st_buffer(dist = 5000) # 5,000 meters

mit_coords <- tibble(lat = 42.360091, lon = -71.09416) |>  # MIT
  st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
  st_transform(26986) |>
  st_buffer(dist = 5000) # 5,000 meters

# Combine the two buffers
combined_vicinity <- st_union(harvard_coords, mit_coords)
```

```{r, message = FALSE, warning = FALSE}
# Filter link volumes to those within 5km of Harvard or MIT

# AM
vicinity_links_am <- link_vols_compare_am |>
  st_transform(26986) |>
  st_filter(combined_vicinity)

vicinity_increases_am <- vicinity_links_am |>
  filter(flow_diff > 50)

vicinity_decreases_am <- vicinity_links_am |>
  filter(flow_diff < -50) |>
  mutate(flow_diff = -1 * flow_diff)

# PM
vicinity_links_pm <- link_vols_compare_pm |>
  st_transform(26986) |>
  st_filter(combined_vicinity)

vicinity_increases_pm <- vicinity_links_pm |>
  filter(flow_diff > 50)

vicinity_decreases_pm <- vicinity_links_pm |>
  filter(flow_diff < -50) |>
  mutate(flow_diff = -1 * flow_diff)

# MD
vicinity_links_md <- link_vols_compare_md |>
  st_transform(26986) |>
  st_filter(combined_vicinity)

vicinity_increases_md <- vicinity_links_md |>
  filter(flow_diff > 50)

vicinity_decreases_md <- vicinity_links_md |>
  filter(flow_diff < -50) |>
  mutate(flow_diff = -1 * flow_diff)

# NT
vicinity_links_nt <- link_vols_compare_nt |>
  st_transform(26986) |>
  st_filter(combined_vicinity)

vicinity_increases_nt <- vicinity_links_nt |>
  filter(flow_diff > 50)

vicinity_decreases_nt <- vicinity_links_nt |>
  filter(flow_diff < -50) |>
  mutate(flow_diff = -1 * flow_diff)
```

# Visualizing comparison of link volumes for each time of day

```{r, message = FALSE, warning = FALSE}
# AM increases
base_map <- get_tiles(vicinity_links_am,
                      provider = "CartoDB.Positron",
                      zoom = 12,
                      crop = TRUE)

ggplot(vicinity_increases_am) +
  geom_spatraster_rgb(data =base_map) +
  geom_sf(aes(color = flow_diff),
          linewidth = 1) +
  scale_color_viridis_c(transform = "log2",
                        breaks = c(80, 160, 320, 640, 1280),
                        direction = -1,
                        name = "Increase in morning\npeak traffic volumes") +
  theme_void()
```

```{r, message = FALSE, warning = FALSE}
# AM decreases
base_map <- get_tiles(vicinity_links_am,
                      provider = "CartoDB.Positron",
                      zoom = 12,
                      crop = TRUE)

ggplot(vicinity_decreases_am) +
  geom_spatraster_rgb(data =base_map) +
  geom_sf(aes(color = flow_diff),
          linewidth = 1) +
  scale_color_viridis_c(transform = "log2",
                        breaks = c(80, 160, 320, 640, 1280, 2560, 5120),
                        direction = -1,
                        name = "Decrease in morning\npeak traffic volumes") +
  theme_void()
```

```{r, message = FALSE, warning = FALSE}
# PM increases
base_map <- get_tiles(vicinity_links_pm,
                      provider = "CartoDB.Positron",
                      zoom = 12,
                      crop = TRUE)

ggplot(vicinity_increases_pm) +
  geom_spatraster_rgb(data =base_map) +
  geom_sf(aes(color = flow_diff),
          linewidth = 1) +
  scale_color_viridis_c(transform = "log2",
                        breaks = c(80, 160, 320, 640, 1280),
                        direction = -1,
                        name = "Increase in afternoon\npeak traffic volumes") +
  theme_void()
```

```{r, message = FALSE, warning = FALSE}
# PM decreases
base_map <- get_tiles(vicinity_links_pm,
                      provider = "CartoDB.Positron",
                      zoom = 12,
                      crop = TRUE)

ggplot(vicinity_decreases_pm) +
  geom_spatraster_rgb(data =base_map) +
  geom_sf(aes(color = flow_diff),
          linewidth = 1) +
  scale_color_viridis_c(transform = "log2",
                        breaks = c(80, 160, 320, 640, 1280, 2560, 5120),
                        direction = -1,
                        name = "Decrease in afternoon\npeak traffic volumes") +
  theme_void()
```

```{r, message = FALSE, warning = FALSE}
# MD increases
base_map <- get_tiles(vicinity_links_md,
                      provider = "CartoDB.Positron",
                      zoom = 12,
                      crop = TRUE)

ggplot(vicinity_increases_md) +
  geom_spatraster_rgb(data =base_map) +
  geom_sf(aes(color = flow_diff),
          linewidth = 1) +
  scale_color_viridis_c(transform = "log2",
                        breaks = c(80, 160, 320, 640, 1280),
                        direction = -1,
                        name = "Increase in mid-day\n off-peak traffic volumes") +
  theme_void()
```

```{r, message = FALSE, warning = FALSE}
# MD decreases
base_map <- get_tiles(vicinity_links_md,
                      provider = "CartoDB.Positron",
                      zoom = 12,
                      crop = TRUE)

ggplot(vicinity_decreases_md) +
  geom_spatraster_rgb(data =base_map) +
  geom_sf(aes(color = flow_diff),
          linewidth = 1) +
  scale_color_viridis_c(transform = "log2",
                        breaks = c(80, 160, 320, 640, 1280, 2560, 5120),
                        direction = -1,
                        name = "Decrease in mid-day\n off-peak traffic volumes") +
  theme_void()
```

```{r, message = FALSE, warning = FALSE}
# NT increases
base_map <- get_tiles(vicinity_links_nt,
                      provider = "CartoDB.Positron",
                      zoom = 12,
                      crop = TRUE)

ggplot(vicinity_increases_nt) +
  geom_spatraster_rgb(data =base_map) +
  geom_sf(aes(color = flow_diff),
          linewidth = 1) +
  scale_color_viridis_c(transform = "log2",
                        breaks = c(80, 160, 320, 640, 1280),
                        direction = -1,
                        name = "Increase in nighttime\n off-peak traffic volumes") +
  theme_void()
```

```{r, message = FALSE, warning = FALSE}
# NT decreases
base_map <- get_tiles(vicinity_links_nt,
                      provider = "CartoDB.Positron",
                      zoom = 12,
                      crop = TRUE)

ggplot(vicinity_decreases_nt) +
  geom_spatraster_rgb(data =base_map) +
  geom_sf(aes(color = flow_diff),
          linewidth = 1) +
  scale_color_viridis_c(transform = "log2",
                        breaks = c(80, 160, 320, 640, 1280, 2560, 5120),
                        direction = -1,
                        name = "Decrease in nighttime\n off-peak traffic volumes") +
  theme_void()
```

# Load transit stop locations

```{r, message = FALSE, warning = FALSE}
file_path = "/Users/paigelee/Documents/SES 5394/T5/stops.geojson"

stop_locs <- st_read(file_path, quiet = TRUE)
```

```{r, message = FALSE, warning = FALSE}
base_map <- get_tiles(stop_locs,
                      provider = "CartoDB.Positron",
                      zoom = 8,
                      crop = TRUE)

ggplot(stop_locs) +
  geom_spatraster_rgb(data = base_map) +
  geom_sf(size = 1, color = "orange") +
  theme_void()
```

# Load boarding/alighting data

We are focusing on all four time periods and only people who walk to/from their transit stops (ignore those who use park and ride lots)

```{r, message = FALSE, warning = FALSE}
# AM

file_path <- "/Users/paigelee/Documents/SES 5394/T5/onoff_tw_am_base.csv"
base_onoff_am <- read_csv(file_path, show_col_types = FALSE) |>
  select(STOP, On, Off) |>
  rename(base_on = On,
         base_off = Off,
         ID = STOP)

file_path <- "/Users/paigelee/Documents/SES 5394/T5/onoff_tw_am_alt.csv"
alt_onoff_am <- read_csv(file_path, show_col_types = FALSE) |>
  select(STOP, On, Off) |>
  rename(alt_on = On,
         alt_off = Off,
         ID = STOP)
```

```{r, message = FALSE, warning = FALSE}
# PM

file_path <- "/Users/paigelee/Documents/SES 5394/T5/onoff_tw_pm_base.csv"
base_onoff_pm <- read_csv(file_path, show_col_types = FALSE) |>
  select(STOP, On, Off) |>
  rename(base_on = On,
         base_off = Off,
         ID = STOP)

file_path <- "/Users/paigelee/Documents/SES 5394/T5/onoff_tw_pm_alt.csv"
alt_onoff_pm <- read_csv(file_path, show_col_types = FALSE) |>
  select(STOP, On, Off) |>
  rename(alt_on = On,
         alt_off = Off,
         ID = STOP)
```

```{r, message = FALSE, warning = FALSE}
# MD

file_path <- "/Users/paigelee/Documents/SES 5394/T5/onoff_tw_md_base.csv"
base_onoff_md <- read_csv(file_path, show_col_types = FALSE) |>
  select(STOP, On, Off) |>
  rename(base_on = On,
         base_off = Off,
         ID = STOP)

file_path <- "/Users/paigelee/Documents/SES 5394/T5/onoff_tw_md_alt.csv"
alt_onoff_md <- read_csv(file_path, show_col_types = FALSE) |>
  select(STOP, On, Off) |>
  rename(alt_on = On,
         alt_off = Off,
         ID = STOP)
```

```{r, message = FALSE, warning = FALSE}
# NT

file_path <- "/Users/paigelee/Documents/SES 5394/T5/onoff_tw_nt_base.csv"
base_onoff_nt <- read_csv(file_path, show_col_types = FALSE) |>
  select(STOP, On, Off) |>
  rename(base_on = On,
         base_off = Off,
         ID = STOP)

file_path <- "/Users/paigelee/Documents/SES 5394/T5/onoff_tw_nt_alt.csv"
alt_onoff_nt <- read_csv(file_path, show_col_types = FALSE) |>
  select(STOP, On, Off) |>
  rename(alt_on = On,
         alt_off = Off,
         ID = STOP)
```

# Compare ridership

### What are the 20 transit stops with the biggest changes in ridership?

```{r, message = FALSE, warning = FALSE}
# AM

stop_diff_am <- stop_locs |>
  inner_join(base_onoff_am) |>
  left_join(alt_onoff_am) |>
  mutate(base_total = base_on + base_off,
         alt_total = alt_on + alt_off,
         diff = alt_total - base_total) |>
  mutate(abs_diff = abs(diff)) |>
  arrange(-abs_diff) |>
  mutate(inc_dec = ifelse(diff < 0, "Decrease", "Increase"))

stop_diff_am |>
  st_drop_geometry() |>
  select(ID, diff) |>
  head(n = 20) |>
  kable()
```

```{r, message = FALSE, warning = FALSE}
# PM

stop_diff_pm <- stop_locs |>
  inner_join(base_onoff_pm) |>
  left_join(alt_onoff_pm) |>
  mutate(base_total = base_on + base_off,
         alt_total = alt_on + alt_off,
         diff = alt_total - base_total) |>
  mutate(abs_diff = abs(diff)) |>
  arrange(-abs_diff) |>
  mutate(inc_dec = ifelse(diff < 0, "Decrease", "Increase"))

stop_diff_pm |>
  st_drop_geometry() |>
  select(ID, diff) |>
  head(n = 20) |>
  kable()
```

```{r, message = FALSE, warning = FALSE}
# MD

stop_diff_md <- stop_locs |>
  inner_join(base_onoff_md) |>
  left_join(alt_onoff_md) |>
  mutate(base_total = base_on + base_off,
         alt_total = alt_on + alt_off,
         diff = alt_total - base_total) |>
  mutate(abs_diff = abs(diff)) |>
  arrange(-abs_diff) |>
  mutate(inc_dec = ifelse(diff < 0, "Decrease", "Increase"))

stop_diff_md |>
  st_drop_geometry() |>
  select(ID, diff) |>
  head(n = 20) |>
  kable()
```

```{r, message = FALSE, warning = FALSE}
# NT

stop_diff_nt <- stop_locs |>
  inner_join(base_onoff_nt) |>
  left_join(alt_onoff_nt) |>
  mutate(base_total = base_on + base_off,
         alt_total = alt_on + alt_off,
         diff = alt_total - base_total) |>
  mutate(abs_diff = abs(diff)) |>
  arrange(-abs_diff) |>
  mutate(inc_dec = ifelse(diff < 0, "Decrease", "Increase"))

stop_diff_nt |>
  st_drop_geometry() |>
  select(ID, diff) |>
  head(n = 20) |>
  kable()
```

### Visualizing the stops with changes of more than n boardings and alightings 

n is determined by the table values (`stop_diff_am`, etc.) in the previous step 

```{r, message = FALSE, warning = FALSE}
# AM

big_stop_diffs_am <- stop_diff_am |>
  filter(abs_diff > 40)

base_map_am <- get_tiles(big_stop_diffs_am,
                      provider = "CartoDB.Positron",
                      zoom = 14,
                      crop = TRUE)

ggplot(big_stop_diffs_am) +
  geom_spatraster_rgb(data = base_map_am) +
  geom_sf(aes(size = abs_diff,
              color = inc_dec),
          alpha = 0.3) +
  scale_size(name = "Magnitude of change") +
  scale_color_manual(name = "Direction of change",
                     values = c("red", "blue"),
                     labels = c("Decrease",
                                "Increase")) +
  labs(title = "Stops with Significant AM Peak Volume Changes") +
  theme_void()
```

```{r, message = FALSE, warning = FALSE}
# PM

big_stop_diffs_pm <- stop_diff_pm |>
  filter(abs_diff > 40)

base_map_pm <- get_tiles(big_stop_diffs_pm,
                      provider = "CartoDB.Positron",
                      zoom = 18,
                      crop = TRUE)

ggplot(big_stop_diffs_pm) +
  geom_spatraster_rgb(data = base_map_pm) +
  geom_sf(aes(size = abs_diff,
              color = inc_dec),
          alpha = 0.3) +
  scale_size(name = "Magnitude of change") +
  scale_color_manual(name = "Direction of change",
                     values = c("red", "blue"),
                     labels = c("Decrease",
                                "Increase")) +
  labs(title = "Stops with Significant PM Peak Volume Changes") +
  theme_void()
```

```{r, message = FALSE, warning = FALSE}
# MD

big_stop_diffs_md <- stop_diff_md |>
  filter(abs_diff > 40)

base_map_md <- get_tiles(big_stop_diffs_md,
                      provider = "CartoDB.Positron",
                      zoom = 15,
                      crop = TRUE)

ggplot(big_stop_diffs_md) +
  geom_spatraster_rgb(data = base_map_md) +
  geom_sf(aes(size = abs_diff,
              color = inc_dec),
          alpha = 0.3) +
  scale_size(name = "Magnitude of change") +
  scale_color_manual(name = "Direction of change",
                     values = c("red", "blue"),
                     labels = c("Decrease",
                                "Increase")) +
  labs(title = "Stops with Significant Mid-Day Off-Peak Volume Changes") +
  theme_void()
```

```{r, message = FALSE, warning = FALSE}
# NT

big_stop_diffs_nt <- stop_diff_nt |>
  filter(abs_diff > 40)

base_map_nt <- get_tiles(big_stop_diffs_nt,
                      provider = "CartoDB.Positron",
                      zoom = 15,
                      crop = TRUE)

ggplot(big_stop_diffs_nt) +
  geom_spatraster_rgb(data = base_map_nt) +
  geom_sf(aes(size = abs_diff,
              color = inc_dec),
          alpha = 0.3) +
  scale_size(name = "Magnitude of change") +
  scale_color_manual(name = "Direction of change",
                     values = c("red", "blue"),
                     labels = c("Decrease",
                                "Increase")) +
  labs(title = "Stops with Significant Nighttime Off-Peak Volume Changes") +
  theme_void()
```
